Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-29 article.tex
proposition de damien pour la A3

As discussed in Section \ref{sec:inference}, the performance of the method (in
terms of AUROC) is sensitive to the value of the parameter $\tau$ of the
hard-threshold filter $h$ (see Equation \ref{eqn:hfilter}), and to the choice
of the low-pass filter (among $\{f_1, f_2, f_3, f_4\}$). Like in the simplified
method, we therefore averaged the partial correlation statistics obtained by
varying $\tau$ linearly from $0.100$ to $0.209$ (in step of $0.001)$ and by
varying the low-pass filter in $\{f_1, f_2, f_3, f_4\}$.

Filters $f_1$ and $f_2$ display similar performances and thus were given equal
weight in the averaging of the simplified method. In contrast, filters $f_3$
and $f_4$ turn out to be individually less competitive and were therefore given
smaller weights in the tuned averaging. Yet, as further shown in
Section~\ref{sapp:results}, combining all $4$ filters proves to slightly
improve performance with respect to using $f_1$ and $f_2$ only.

@@

Partial correlation statistics is a symmetric measure, while the connectome is
a directed graph. It could thus be beneficial to try to predict edge orientation. In
this section, we present a heuristic to retrieve directionality. %% Note however that
%% given the discussion at the end of Section~\ref{sec:results}, we cannot expect
%% a very large improvement in AUROC by predicting edge orientation.

This heuristic is based on the following observation. The rise of fluorescence
of a neuron indicates its activation. If another neuron is activated after a
slight delay, this could be a consequence of the activation of the first neuron
and therefore indicate a directed link in the connectome from the first to the
second neuron. Given this analysis, the idea of our heuristic is to count the
number of such events occurring between neurons $i$ and $j$ for consecutive
time steps. Mathematically, we compute for each pair $(i,j)$:
\begin{align}
s_{i,j} = \sum_{t=1}^{T - 1} \mathbb{1}((x_j^{t+1} - x_i^t) \in \left[\phi_1, \phi_2\right])
\end{align}
where $\phi_1$ and $\phi_2$ are parameters of the heuristic (with $\phi_1 = 0.2$ and $\phi_2 = 0.5$ in our experiments).

Two issues remain with this heuristic. First, it does not allow to detect a
directed relationship if both neurons are activated simultaneously at the same
time step, i.e., when the delay between the two activations is below the time
resolution. Second, one needs to compare $s_{j,i}$ and $s_{i,j}$ to decide on
edge orientation. Indeed, edge orientation can only be predicted with some
confidence if either $s_{i,j}$ is greater than $s_{j,i}$ ($i\rightarrow j$) or
$s_{i,j}$ is lower than $s_{j,i}$ ($j\rightarrow i$). To address these issues,
we computed a differential pairwise coefficient $z_{i,j}$ defined as follows:
\begin{align}
z_{i,j} = s_{i,j} - s_{j,i}.
\end{align}
The coefficient $z_{i,j}$ estimates the confidence in an edge $i\rightarrow j$. Note that if $z_{i,j}$ is positive, then $z_{j,i}$ will be negative.

This heuristic does not give good results on its own (see Table \ref{tab:directivity} in Section~\ref{sapp:results}). However, it can be used to asymmetrize the degree of association (symmetric) matrix given by partial correlation statistics. Concretely, a new matrix is computed as the linear combination of the partial correlation matrix and the heuristic matrix.
Mathematically, we have
\begin{align}
q_{i,j} = weight_p * p_{i,j} + weight_z * z_{i,j}
\end{align}
where $weight_p$ and $weight_z$ are parameters of the method (with $weight_p = 0.997$ and $weight_z = 0.003$ in our experiments).

Considering the small weight associated to this heuristic, the effect is subtle. It slightly increases the confidence in one edge direction and decreases the confidence in the opposite edge direction while not affecting strongly the undirected topology of the network. This process slightly improves the performance of our method (see Table \ref{tab:directivity} in Section~\ref{sapp:results}).

>>>>

As discussed in Section \ref{sec:inference}, the performance of the method (in
terms of AUROC) is sensitive to the value of the parameter $\tau$ of the
hard-threshold filter $h$ (see Equation \ref{eqn:hfilter}), and to the choice
of the low-pass filter (among $\{f_1, f_2, f_3, f_4\}$).
Like in the simplified method, we have averaged the partial correlation statistics obtained for all the pairs $(\tau,\mbox{low-pass filter}) \in \{0.100,0.101,\ldots,0.209\}\times \{f_1, f_2, f_3, f_4\}$.

Filters $f_1$ and $f_2$ display similar performances and thus were given equal
weight. Additionally, this weight were chosen equal as the weight selected for the simplified method. In contrast, filters $f_3$
and $f_4$ turn out to be individually less competitive and were therefore given
smaller weights in the tuned averaging. Yet, as further shown in
Section~\ref{sapp:results}, combining all $4$ filters proves to slightly
improve performance with respect to using $f_1$ and $f_2$ only.

@@

Partial  correlation  statistics is  a  symmetric  measure, while  the
connectome is a directed graph. It  could thus be beneficial to try to
predict edge orientation. In this section, we present a heuristic that
modifies the  $p_{ij}$ computed  by the  approach described  before to
take into account directionality.

This approach is based on the following
observation. The rise of fluorescence of a neuron indicates its
activation. If another neuron is activated after a slight delay, this
could be a consequence of the activation of the first neuron and
therefore indicates a directed link in the connectome from the first to
the second neuron.  Given this observation, we have computed for every
pair $(i,j)$ an empirical estimation of the  term:
\begin{align}
s_{i,j} = \sum_x \sum_{t=1}^{T - 1} \mathbb{1}((x_j^{t+1} - x_i^t) \in \left[\phi_1, \phi_2\right])
\end{align}
that could be interpreted as an image of the  number of times
that neuron $i$ activates neuron $j$. $\phi_1$ and $\phi_2$ are
parameters whose values have been chosen in our experiments equal to   
$0.2$ and $0.5$, respectively. Their role is
define when the difference between $x_j^{t+1}$  and $x_i^t$ can
indeed be assimilated to an event where neuron $i$ activates neuron
$j$.

Afterwards, we have computed the difference between $s_{i,j}$ and
$s_{j,i}$, that we call $z_{i,j}$, and used this difference to modify  $p_{i,j}$ and
$p_{i,j}$ so as to take into account directionality. Naturally, îf
$z_{i,j}$ is greater  (smaller) than $0$, we may conclude that should there  be an
edge between $i$ and $j$, then this edge would have to be oriented
from $i$ to $j$ ($j$ to $i$).

This suggests the the new association matrix $r$:
\begin{align}
r_{i,j} =  \mathbb{1}(z_{i,j} > \phi_3)    p_{i,j}
\end{align}
where $\phi_3 >0$ is another parameter. We found out that this new
matrix $r$ was not giving good results, probably due to the fact that
directivity was not well enough rewarded in the challenge.

This has lead us to investigate other ways for exploiting the
information about directionality contained in the matrix $z$. One of
those ways that gave good performance was to use as association
matrix: 
\begin{align}
q_{i,j} = weight * p_{i,j} + (1-weight) * z_{i,j}
\end{align}
with  $weight$ chosen close to 1 ($weight=0.997$). Note that with
values for $weight$ close to 1,   matrix $q$ only slightly uses the
information about directivity contained in $z$ to modify the  partial
correlation matrix $p$. We tried smaller values for $weight$ but those
were leading to worse results.

It was  this association matrix $q_{i,j}$ that actually led to the
best results of the challenge, as shown in Table \ref{tab:directivity}
of Section~\ref{sapp:results}.
